---
title: "Peer Assessment #1 - Reproducible Research"
output: 
  html_document: 
    keep_md: yes
---

**Peer assignment #1 for the Reproducible Research course offered by Coursera**

### Loading and preprocessing the data

---

1. Load the data

*This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.*

The dataset is already present in the working folder as a plaintext CSV file.

Source: Forked from https://github.com/rdpeng/RepData_PeerAssessment1 on 20/09/2015

Variables included in this dataset:

- **steps**: Number of steps taken in a 5-minute interval (NA represents missing values) 
- **date**: The date the measurement was taken in YYYY-MM-DD format 
- **interval**: Identifier for the 5-minute interval in which measurement was taken

* Load the data into a dataframe and inspect the structure of the data
```{r}
dat = read.csv('activity.csv', header = T)
names(dat)
str(dat)
head(dat)
```


### What is the mean total number of steps taken per day?

---

For this part of the assignment, we can ignore missing values.

We need to:
1. Calculate the total number of steps taken per day,
2. Make a histogram of the total number of steps taken each day,
3. Calculate and report the mean and median of the total number of steps taken per day.

1. Calculate the total number of steps taken per day
```{r}
library(data.table)
dat_tbl = data.table(dat)
dat_tbl_summary = dat_tbl[, list(total_steps = sum(steps, na.rm = T)), 
                          by = date]
```

The data frame dat_tbl_summary contains the required data. The first few lines are provided below as a preview.
```{r}
head(dat_tbl_summary)
```

2. Make a histogram of the total number of steps taken each day

**Note: Mean and Median are reported in the Histogram**

```{r}
# Function to generate the histogram
gen_hist = function(x, title){
        hist(x, 
             breaks = 20,
             main = title,
             xlab = 'Total Number of Steps', col = 'grey',
            
             cex.main = .9)
        
        #caluclate mean and median
        mean_value = round(mean(x), 2)
        median_value = round(median(x), 2)
        
        #place lines for mean and median on histogram
        abline(v=mean_value, lwd = 3, col = 'blue')
        abline(v=median_value, lwd = 3, col = 'red')
        
        #create legend
        legend('topright', lty = 1, lwd = 3, col = c("blue", "red"),
               cex = .8, 
               legend = c(paste('Mean: ', mean_value),
               paste('Median: ', median_value))
               )
}

gen_hist(dat_tbl_summary$total_steps, 'Number of Steps Taken Per Day')
```

3. Calculate and report the mean and median of the total number of steps taken per day
```{r}
mean(dat_tbl_summary$total_steps)
median(dat_tbl_summary$total_steps)
```


### What is the average daily activity pattern?

---

1.  Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
2.  Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?


```{r}
# Summarize the dataset by interval
dat_tbl_summary_intv = dat_tbl[, list(avg_steps = mean(steps, na.rm = T)), 
                          by = interval]
# Plot the time series
with(dat_tbl_summary_intv, {
        plot(interval, avg_steps, type = 'l',
             main = 'Average Steps by Time Interval',
             xlab = '5 Minute Time Interval',
             ylab = 'Average Number of Steps')
        })
# Find the interval that has the maximum avg steps
max_steps = dat_tbl_summary_intv[which.max(avg_steps), ]

# Generate label string
max_lab = paste('Maximum Of ', round(max_steps$avg_steps, 1), ' Steps \n On ', max_steps$interval, 'the Time Interval', sep = '')

# Collect co-oridinates of the max interval
points(max_steps$interval,  max_steps$avg_steps, col = 'red', lwd = 3, pch = 19)

# Annotate maximum # steps And interval
legend("topright",
       legend = max_lab,
       text.col = 'red',
       bty = 'n'
       )

```

---

### Imputing missing values

1. Calculate & Report The Number of Missing Values
```{r}
sum(is.na(dat$steps))
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
# Join the dataframe created earlier that summarizes the average number of steps per interval to the original dataset
setkey(dat_tbl, interval)
setkey(dat_tbl_summary_intv, interval)


# Create a function that will return the second value if the first value is NA
NA_replace = function(x,y){
        if(is.na(x)){
                
                return(y)
        }
        return(x)
}

# Create a new dataset that replaces NAs with average values
dat_tbl_miss = dat_tbl[dat_tbl_summary_intv]
dat_tbl_miss$new_steps = mapply(NA_replace,dat_tbl_miss$steps, dat_tbl_miss$avg_steps)

# Summarize the new dataset by day
dat_tbl_summary_miss = dat_tbl_miss[, list(new_steps = sum(new_steps, na.rm = T)), 
                          by = date]
# Preview the new dataset
head(dat_tbl_summary_miss)
```

4.  Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

**Note: Mean and Median are reported in the Histogram**

```{r}

gen_hist(dat_tbl_summary$total_steps, 'Missing Values Removed')
gen_hist(dat_tbl_summary_miss$new_steps, 'Missing Values Replaced With \n Mean For Interval')

```

The mean and the median are now almost the same after replacing missing values with the mean value for the relevant interval. It makes sense that the median value would now move closer to the mean. So the Median value increased after this method of missing value replacement.

### Are there differences in activity patterns between weekdays and weekends?

---

1.  Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
# Make a function to return either "Weekday" or "Weekend"
weekpart = function(x){
        if(x %in% c('Saturday', 'Sunday')){
                return('Weekend')
        }
        
        return('Weekday')
}

# Add name of week
dat_tbl_miss$dayname = weekdays(as.Date(dat_tbl_miss$date))

# Add factor variable to differentiate Weekdays and Weekends
dat_tbl_miss$daytype = as.factor(apply(as.matrix(dat_tbl_miss$dayname), 1, weekpart))

# Summarize dataset: Mean grouped by interval and daytype
dat_tbl_summary_miss = dat_tbl_miss[, list(avg_steps = mean(new_steps, na.rm = T)), 
                          by = list(interval, daytype)]

# Inspect dataset
str(dat_tbl_summary_miss)
```

Below is the panel plot:
```{r}
library(lattice)
xyplot(avg_steps~interval | daytype, data = dat_tbl_summary_miss,
      type = 'l',
      xlab = 'Interval',
      ylab = 'Number of Steps',
      layout = c(1,2))
```
